Teaching Low-Level Concurrency with \pebbles and \landslide

### 2016-08-27 sigcse-17/410.tex
sigcse: garth's revisions

\fourten, Operating System Design and Implementation,
is an upper-level undergraduate class
designed for juniors and seniors in Computer Science
and Electrical and Computer Engineering.
For CS students it is one of five courses which fulfill
a ``systems programming'' degree requirement.
%
The prerequisite course is \twothirteen,
Introduction to Computer Systems,
% TODO CAMREADY
%a course which has been adopted at other
%universities in the U.S.\ and worldwide
adapted from Carnegie Mellon University's 15-213
\cite{sigcse01:CSaPP}.
This section discusses the course learning objectives,
outlines the programming projects,
and mentions some current limitations in the scope
of the course material.

@@

The first project is a backtrace library: %``stack crawler'':
when invoked by a client program, it displays the
%program's
stack symbolically,
showing function names, strings, floating-point values, etc.
This project
%\shortversion{enables}{meets a variety of goals: it enables}
%students to review key process-model and
%language-runtime concepts from the
%prerequisite course;
%it
introduces students to our expectations about
design, analysis, and making choices;
%finally,
and
because C pointers are unsafe, it requires students
to consider robustness.
%
% and we READ THEIR CODE
% ^^ LOL, classic dave.
This project is built using a standard C tool chain
and debugged as students see fit (generally \x{printf()} and GDB).

@@

This project and the remaining ones are written in
C with some x86-32 assembly,
which is then
compiled and linked into an ELF executable,
stored into a %1.44-megabyte 3.5-inch
floppy-disk image,
and booted via GRUB.
If the image is copied to a real floppy or CD,
%or \shortversion{compact disc,}{embedded into an ``El Torito'' bootable compact disc image,}
it can boot on standard PC hardware.
However, students most often use
Simics \cite{simics} (described in \sect{\ref{sec:simics}}).

@@

The third project is implementing a 1:1 thread library for
user-space programs,
%essentially a stripped-down
%version of POSIX Pthreads.
similar to POSIX Pthreads.
The primary goal is
for students to experiment
with concurrency and atomicity.
Students begin by designing mutexes using any
x86-32 atomic instructions they choose.
%(there is substantial variation across groups).
They then write other synchronization
primitives (condition variables, semaphores,
and reader/writer locks), infrastructure
components (stack allocation/recycling and
a thread registry),
and low-level code to launch and shut down
threads.

@@

Student library code is linked against small
test programs provided by the course staff,
%producing one statically linked ELF executable
%per test program.}
%The test programs \shortversion{run on}{are bundled into a RAM-disk
%image and linked against}
which then run on a ``reference kernel'' provided in binary form.
%written by
%the course staff
%and provided in binary form.
The reference kernel's behavior is specified
% TODO CAMREADY: deanonymize
in a twelve-page document~\cite{kspec-anonymized}.
In addition to providing a reliable execution
substrate,
the reference kernel
schedules user-space threads created by
student code using a variety of
interleaving policies.
%\shortversion{}{In order to aid debugging of concurrency problems,
%the \x{misbehave()} system call enables
%students to manually select an interleaving policy.}

@@

For the fourth project, students %two-student teams
produce a kernel which implements the
same specification as the reference kernel
they previously relied on.
They design and implement some approach to
synchronizing and blocking threads while
they are in kernel space,
%\shortversion{,}{\footnote{
%Student designs for their kernel's synchronization primitives vary widely.
%For example, ``top of the class'' students who went on to become teaching assistants have alternately used condition variables but not semaphores, and semaphores but not condition variables.
%},}
a simple round-robin scheduler,
basic virtual memory,
a program loader,
%various x86 exception handlers,
and code for setting up and tearing down
threads and processes
(re-using their game-project device drivers).
These ingredients are combined to yield
twenty-five system calls,
which we list in Table~\ref{tab:syscalls}.
%(life-cycle, thread management, memory management,
%console I/O, and miscellaneous),
%which we discuss further in Section~\ref{sec:pebbles}.}

\begin{table}[t]
	\center \footnotesize
	\begin{tabular}{|l|p{0.30\textwidth}|}
		\hline
		\bf Syscall name & \bf Summary \\
		\hline
		\multicolumn{2}{c}{\em Lifecycle management} \\
		\hline
		\x{fork} & Duplicates the invoking task, including all memory regions. \\
		\x{thread_fork} & Creates a thread in the current task.\\
		\x{exec} & Replaces the current program in the invoking task with a new one. \\
		\x{set_status} & Records exit status of current task. \\
		\x{vanish} & Ends execution of the calling thread. \\
		\x{wait} & Blocks execution until another task terminates; collects its exit status.\\
		\x{task_vanish}* & Causes all threads of a task to \x{vanish}. \\
		\hline
		\multicolumn{2}{c}{\em Thread management} \\
		\hline
		\x{gettid} & Returns the ID of the invoking thread. \\
		\x{yield} & Defers execution to a specified thread. \\
		\x{deschedule} & Blocks execution of invoking thread. \\
		\x{make_runnable} & Wakes another \x{deschedule}d thread. \\
		\x{get_ticks} & Gets timer tick count since bootup. \\
		\x{sleep} & Blocks for a given number of ticks. \\
		\x{swexn} & Registers a user-space function as a software exception handler.\\
		\hline
		\multicolumn{2}{c}{\em Memory management} \\
		\hline
		\x{new_pages} & Allocates a specified memory region. \\
		\x{remove_pages} & Deallocates same. \\
		\hline
		\multicolumn{2}{c}{\em Console I/O} \\
		\hline
		\x{getchar}* & Reads one character from kbd input. \\
		\x{readline} & Reads the next line from kbd input. \\
		\x{print} & Prints a memory buffer to the console. \\
		\x{set_term_color} & Sets color for future console output. \\
		\x{set_cursor_pos} & Sets the console cursor location. \\
		\x{get_cursor_pos} & Retrieves the console cursor location. \\
		\hline
		\multicolumn{2}{c}{\em Miscellaneous} \\
		\hline
		\x{ls} & Loads a buffer with the names of files stored in the RAM disk ``file system.'' \\
		\x{halt} & Ceases execution of the OS. \\
		\x{misbehave}* & Selects thread-scheduling policy. \\
		\hline
	\end{tabular}
	\caption{The \pebbles specifcation defines 25 system calls. Students are not required to implement ones marked with an asterisk (*), though the reference kernel provides them. }
	\label{tab:syscalls}
\end{table}

@@

Widely regarded as the most difficult concurrency problem in the project
is that of coordinating a parent and a child task that ``simultaneously''
exit:
% invoke \x{vanish()}:
when a task completes,
live children and exited zombies must be handed off
to the task's parent or to the system's \x{init} process,
at a time when the task's parent may itself be
exiting;
meanwhile, threads in tasks that receive new children
may need to be awakened from the \x{wait()} system call.
Due to design constraints imposed by other parts of the kernel specification,
careless solutions %solutions which are not carefully designed
are prone to data races or deadlocks.

@@

Students who complete the kernel project on time
then work on a kernel-extension project.
Past projects have included
a sound card driver,
a filesystem,
suspend-to-disk,
profiling,
and an in-kernel debugger.
Two recent, more aggressive, projects have been
paravirtualization
%so that their kernels can host guest kernels and
and multi-processor support.
%to their single-processor kernels.

@@

\textbf{Automated tests.}
%While developing their code,
Students will uncover both bugs
and design misunderstandings by running test code we
provide.
%Some tests are simple, exercising one feature or condition
%(e.g., allocating and freeing a large amount of memory),
%while other tests exercise a mix of kernel features
%simultaneously (creating and destroying processes,
%some of which do I/O while others handle or are killed
%by various exceptions) over an extended period of time.
The test suite is a number of user-space programs
which test
%functionality including
basic thread and process lifecycles, robustness
against corner-case security vulnerabilities, and long-term stability
under large stress tests.
%
%When students submit their thread libraries and kernels,
%we use an automated testing harness to
%run a superset of the public test suite.
%
Of course, due to the nondeterministic nature of many bugs in student code,
these tests can provide only a rough estimate of a submission's correctness,
rather than verifying it with certainty.
(This hit-or-miss nature motivated our work on \landslide.)
%Nevertheless, performance on these tests constitutes 30\% of the kernel project's grade.

\textbf{Hurdle completion criteria.}
Because of the ambitious learning objectives,
we employ a very detailed notion of completeness
for the thread library and kernel projects.
The ``hurdle'' model is based on the following observations:

@@

The thread library hurdle requires students to pass a suite of five test cases,
which test mutex locking/unlocking and thread creation, exit, and joining in moderately stressful ways.
%
A hurdle-passing kernel can boot up and run the shell,
can pass 80\% of the basic system-call sanity test,
70\% of the solidity tests (which
invoke system calls in illegal
%or non-trivial
ways),
and two of the three long-running multi-feature tests.
%In addition,
%a complete kernel is certified by the authors to
Students also certify their thread libraries and kernels to
be free of ``concurrency shortcuts,''
to handle resource exhaustion,
and to avoid certain distasteful x86-specific code structures.

>>>>

\fourten, Operating System Design and Implementation,
is an upper-level undergraduate class
designed for juniors and seniors in Computer Science
% garth removed this, i guess
%and Electrical
and Computer Engineering.
For CS students it is one of five courses which fulfill
a ``systems programming'' degree requirement.
%
The prerequisite course is \twothirteen,
Introduction to Computer Systems,
% TODO CAMREADY
%a course which has been adopted at other
%universities in the U.S.\ and worldwide
adapted from Carnegie Mellon University's 15-213
\cite{sigcse01:CSaPP}.
This section discusses the course learning objectives,
outlines the programming projects,
and mentions some current limitations in the scope
of the course material.

@@

The first project is a backtrace library: %``stack crawler'':
when invoked by a client program, it displays the
%program's
stack symbolically,
showing function names, strings, floating-point values, etc.
This project
%\shortversion{enables}{meets a variety of goals: it enables}
%students to review key process-model and
%language-runtime concepts from the
%prerequisite course;
%it
introduces students to course expectations about
design, analysis, and making choices;
%finally,
and
because C pointers are unsafe, it requires students
to consider robustness.
%
% and we READ THEIR CODE
% ^^ LOL, classic dave.
This project is built using a standard C tool chain
% FIXME garth wants to cite gdb
and debugged as students see fit (generally \x{printf()} and GDB).

@@

This project and the remaining ones are written in
C with a small amount of x86-32 assembly code,
which is then
compiled and linked into an ELF executable,
stored into a %1.44-megabyte 3.5-inch
floppy-disk image,
% FIXME garth wants to cite grub
and booted via GRUB.
If the image is copied to a real floppy or CD,
%or \shortversion{compact disc,}{embedded into an ``El Torito'' bootable compact disc image,}
it can boot on standard PC hardware.
However, students most often use
the x86 hardware simulator
Simics \cite{simics} (described in \sect{\ref{sec:simics}}).

@@

The third project is implementing a 1:1 thread library for
user-space programs,
%essentially a stripped-down
%version of POSIX Pthreads.
% FIXME garth wants to cite pthreads
similar to POSIX Pthreads.
The primary goal is
for students to experiment
with concurrency and atomicity.
Students begin by designing mutexes using any
x86-32 atomic instructions they choose.
%(there is substantial variation across groups).
They then write other synchronization
primitives (condition variables, semaphores,
and reader/writer locks), infrastructure
components (stack allocation/recycling and
a thread registry),
and low-level code to launch and shut down
threads.

@@

Student library code is linked against small
test programs provided by the course staff,
%producing one statically linked ELF executable
%per test program.}
%The test programs \shortversion{run on}{are bundled into a RAM-disk
%image and linked against}
which then run on a ``reference kernel'' provided in binary form.
%written by
%the course staff
%and provided in binary form.
The reference kernel's behavior is specified
% TODO CAMREADY: deanonymize
in a twelve-page document~\cite{kspec-anonymized}.
In addition to providing a reliable execution
substrate,
the reference kernel
schedules user-space threads created by
student code using a variety of
interleaving policies,
with the ad-hoc goal of exposing concurrency bugs.
%\shortversion{}{In order to aid debugging of concurrency problems,
%the \x{misbehave()} system call enables
%students to manually select an interleaving policy.}

@@

For the fourth project, students %two-student teams
produce a kernel which implements the
same specification as the reference kernel
they previously relied on.
They design and implement some approach to
synchronizing and blocking threads while
they are in kernel space,
%\shortversion{,}{\footnote{
%Student designs for their kernel's synchronization primitives vary widely.
%For example, ``top of the class'' students who went on to become teaching assistants have alternately used condition variables but not semaphores, and semaphores but not condition variables.
%},}
a simple round-robin scheduler,
basic virtual memory,
a program loader,
%various x86 exception handlers,
and code for setting up and tearing down
threads and processes
(re-using their game-project device drivers).
These ingredients are combined to yield
twenty-five system calls,
listed in Table~\ref{tab:syscalls}.
%(life-cycle, thread management, memory management,
%console I/O, and miscellaneous),
%which we discuss further in Section~\ref{sec:pebbles}.}

\begin{table}[t]
	\center \footnotesize
	\begin{tabular}{|l|p{0.30\textwidth}|}
		\hline
		\bf Syscall name & \bf Summary \\
		\hline
		\multicolumn{2}{c}{\em Lifecycle management} \\
		\hline
		\x{fork} & Duplicates the invoking task, including all memory regions. \\
		\x{thread_fork} & Creates a thread in the current task.\\
		\x{exec} & Replaces the current program in the invoking task with a new one. \\
		\x{set_status} & Records exit status of current task. \\
		\x{vanish} & Ends execution of the calling thread. \\
		\x{wait} & Blocks execution until another task terminates; collects its exit status.\\
		\x{task_vanish}* & Causes all threads of a task to \x{vanish}. \\
		\hline
		\multicolumn{2}{c}{\em Thread management} \\
		\hline
		\x{gettid} & Returns the ID of the invoking thread. \\
		\x{yield} & Defers execution to a specified thread. \\
		\x{deschedule} & Blocks execution of invoking thread. \\
		\x{make_runnable} & Wakes another \x{deschedule}d thread. \\
		\x{get_ticks} & Gets timer tick count since bootup. \\
		\x{sleep} & Blocks for a given number of ticks. \\
		\x{swexn} & Registers a user-space function as a software exception handler.\\
		\hline
		\multicolumn{2}{c}{\em Memory management} \\
		\hline
		\x{new_pages} & Allocates a specified memory region. \\
		\x{remove_pages} & Deallocates same. \\
		\hline
		\multicolumn{2}{c}{\em Console I/O} \\
		\hline
		\x{getchar}* & Reads one character from keyboard. \\
		\x{readline} & Reads the next line from keyboard. \\
		\x{print} & Prints a memory buffer to the console. \\
		\x{set_term_color} & Sets color for future console output. \\
		\x{set_cursor_pos} & Sets the console cursor location. \\
		\x{get_cursor_pos} & Retrieves the console cursor location. \\
		\hline
		\multicolumn{2}{c}{\em Miscellaneous} \\
		\hline
		\x{ls} & Loads a buffer with the names of files stored in the RAM disk ``file system.'' \\
		\x{halt} & Ceases execution of the OS. \\
		\x{misbehave}* & Selects thread-scheduling policy. \\
		\hline
	\end{tabular}
	\caption{The \pebbles specifcation defines 25 system calls. Students are not required to implement calls marked with an asterisk (*), though the reference kernel provides them. }
	\label{tab:syscalls}
\end{table}

@@

%Widely regarded as
The most difficult concurrency problem in the project
is that of coordinating a parent and a child task that ``simultaneously''
exit:
% invoke \x{vanish()}:
when a task completes,
live children and exited zombies must be handed off
to the task's parent or to the system's \x{init} process,
at a time when the task's parent may itself be
exiting;
meanwhile, threads in tasks that receive new children
may need to be awakened from the \x{wait()} system call.
Due to design constraints imposed by other parts of the kernel specification,
careless solutions %solutions which are not carefully designed
are prone to data races or deadlocks.

@@

Students who complete the kernel project on time
then work on a kernel-extension project.
Past projects have included
a sound card driver,
a filesystem,
suspend-to-disk,
profiling,
and an in-kernel debugger.
Two more aggressive, projects have been
paravirtualization
%so that their kernels can host guest kernels and
and multi-processor support.
%to their single-processor kernels.

@@

\textbf{Automated tests.}
%While developing their code,
Students uncover both bugs
and design misunderstandings by running course-provided test code.
%Some tests are simple, exercising one feature or condition
%(e.g., allocating and freeing a large amount of memory),
%while other tests exercise a mix of kernel features
%simultaneously (creating and destroying processes,
%some of which do I/O while others handle or are killed
%by various exceptions) over an extended period of time.
The test suite is a number of user-space programs
which test
%functionality including
basic thread and process lifecycles, robustness
against corner-case security vulnerabilities, and long-term stability
under long-running stress tests.
%
%When students submit their thread libraries and kernels,
%we use an automated testing harness to
%run a superset of the public test suite.
%
Of course, due to the nondeterministic nature of many bugs in student code,
these tests can provide only a rough estimate of a submission's correctness,
rather than verifying it with certainty.
(This hit-or-miss nature motivated our work on \landslide.)
%Nevertheless, performance on these tests constitutes 30\% of the kernel project's grade.

\textbf{Hurdle completion criteria.}
Because of the ambitious learning objectives,
we employ a detailed notion of completeness
for the thread library and kernel projects.
The ``hurdle'' model is based on the following observations:

@@

The thread library hurdle requires students to pass a suite of five test cases,
which test mutex locking/unlocking and thread creation, exit, and joining in moderately stressful ways.
%
A hurdle-passing kernel can boot up and run the shell,
can pass 80\% of the basic system-call sanity test,
70\% of the solidity tests (which
invoke system calls in illegal
%or non-trivial
ways),
and two of the three long-running multi-feature tests.
%In addition,
%a complete kernel is certified by the authors to
Students also certify their thread libraries and kernels to
be free of ``concurrency shortcuts,''
to handle resource exhaustion,
and to avoid certain prohibited x86-specific code structures.

### 2016-08-27 sigcse-17/eval.tex
sigcse: garth's revisions

We made \landslide available to students for the thread library project
during the Spring '15, Fall '15, and Spring '16 semesters of \fourten.
We introduced stateless model checking and gave a demo during a special lecture,
and required students to pass the thread library hurdle (\sect{\ref{sec:grading}}) to use \landslide.
Of 90 total two-student groups in those semesters, 47 tested their thread libraries with \landslide.

@@

\begin{table}[t]
	\begin{tabular}{r|cc|cc}
	%det bugs found	det bugs fixed	races fixed	races found
	& \multicolumn{2}{c|}{Deterministic bugs} & \multicolumn{2}{c}{Nondeterministic bugs} \\
	\# bugs	& found & fixed & found & fixed \\
	\hline
	0	& 27	& 32	& 15	& 23	\\
	1	& 6	& 5	& 7	& 11	\\
	2	& 9	& 6	& 12	& 8	\\
	3	& 2	& 1	& 7	& 2	\\
	4	& 1	& 1	& 3	& 1	\\
	5	& 2	& 2	& 2	& 1	\\
	11	& 	& 	& 1	& 1	\\
	\hline
	%Total groups
	%	& 47	& 47	& 47	& 47	\\
	Total bugs
		& 44	& 34	& 85	& 53	\\
	\end{tabular}
	\caption{Summary of how many bugs were found and/or fixed by how many groups.
	Each column counts how many groups found/fixed the number of bugs in each row;
	for instance, two groups found 5 concurrency bugs, one of which fixed all 5.}
	\label{tab:this-table-sucks-but-it's-the-best-i-got}
\end{table}

@@

\begin{figure}[t]
	\includegraphics[width=0.48\textwidth]{p2-distribution-large.pdf}
	\caption{Distribution of thread library grades.}
	\label{fig:p2-distribution}
\end{figure}
We also collected the overall thread library grades to evaluate whether \landslide helps students achieve more during the project.
Figure~\ref{fig:p2-distribution} shows the distribution of these grades.
Compared to students from the same semesters who opted not to use \landslide,
those who did scored on average 3\% higher. % that's rough, buddy.
%
%However, to account for the possible bias for opting-out students being more at the bottom of the class
%on account of not enough free time to volunteer for \landslide,
However, the latter distribution is potentially biased towards struggling students
who lack enough free time to volunteer for \landslide.
To account for this, we also checked grades from prior semesters when \landslide was not available.
%This distribution is indistiguishable from the experimental group,
Although the minimum grade among the experimental group is 17\% higher than the minimum from prior semesters,
the middle 50\% of the distributions are quite similar,
suggesting that \landslide's benefit is largely offset by the time investment required.
%
In future work, we will streamline the user experience to help students fix as many concurrency bugs with a smaller time investment.

>>>>

We made \landslide available to students for the thread library project
during the Spring '15, Fall '15, and Spring '16 semesters of \fourten.
We introduced stateless model checking in class, %and gave a demo during a special lecture,
and required students to pass the thread library hurdle (\sect{\ref{sec:grading}}) to use \landslide.
Of 90 two-student groups in those semesters, 47 tested their thread libraries with \landslide.

@@

\begin{table}[t]
	\begin{tabular}{r|cc|cc}
	%det bugs found	det bugs fixed	races fixed	races found
	& \multicolumn{4}{c}{Number of groups} \\
	Number & \multicolumn{2}{c|}{Deterministic bugs} & \multicolumn{2}{c}{Nondeterministic bugs} \\
	of bugs	& found & fixed & found & fixed \\
	\hline
	0	& 27	& 32	& 15	& 23	\\
	1	& 6	& 5	& 7	& 11	\\
	2	& 9	& 6	& 12	& 8	\\
	3	& 2	& 1	& 7	& 2	\\
	4	& 1	& 1	& 3	& 1	\\
	5	& 2	& 2	& 2	& 1	\\
	11	& 	& 	& 1	& 1	\\
	\hline
	%Total groups
	%	& 47	& 47	& 47	& 47	\\
	Total bugs
		& 44	& 34	& 85	& 53	\\
	\end{tabular}
	\caption{Summary of how many bugs were found and/or fixed by how many groups.
	Each column counts how many groups found/fixed the number of bugs in each row;
	for instance, two groups found 5 concurrency bugs, one of which fixed all 5.}
	\label{tab:this-table-sucks-but-it's-the-best-i-got}
\end{table}

@@

\begin{figure}[t]
	\includegraphics[width=0.48\textwidth]{p2-distribution-logos.pdf}
	\caption{Distribution of thread library grades.}
	\label{fig:p2-distribution}
\end{figure}
We also collected the overall thread library grades to evaluate whether \landslide helps students achieve more during the project.
Figure~\ref{fig:p2-distribution} shows the distribution of these grades.
Compared to students from the same semesters who opted not to use \landslide,
those who did scored on average 3\% higher. % that's rough, buddy.
%
%However, to account for the possible bias for opting-out students being more at the bottom of the class
%on account of not enough free time to volunteer for \landslide,
However, the latter distribution is potentially biased towards struggling students
who lack enough free time to volunteer for \landslide.
To account for this, we also checked grades from prior semesters when \landslide was not available.
%This distribution is indistiguishable from the experimental group,
Although the minimum grade among the experimental group is 17\% higher than the minimum from prior semesters,
the middle 50\% of the distributions are quite similar,
suggesting that \landslide's benefit is largely offset by the time investment required.
%
In future work, we will streamline the user experience to help students fix as many concurrency bugs with a smaller time investment.

### 2016-08-27 sigcse-17/intro.tex
sigcse: rename double blind name of pebbles/landslide
sigcse: garth's revisions

In this paper we describe our experience teaching concurrent programming at the operating-system level to advanced undergraduate students at \carnegiemellon (\cmu).
In \fourten, the Operating Systems Design and Implementation class at \cmu,
students complete aggressively low-level thread library and kernel projects,
which run on real hardware,
in which they implement threads and processes.
Presented with a kernel specification called \pebbles
% TODO camready
\cite{kspec-anonymized}
, students build a user-space threading and synchronization library on top of a \pebbles-compliant kernel in a two-week project, and subsequently, in a six-week project, implement their own multi-threaded kernel.
%During these projects
The students face demanding synchronization problems, including how to implement locking primitives and how to coordinate the possibly-simultaneous exiting of parent and child processes.
% Finally, in another two-week project, students implement an extension to their Pebbles kernel. This project changes every semester, and recently has involved supporting symmetric multiprocessing (SMP).

In comparison with other educational kernel projects,
\pebbles focuses more on ground-up design than on teaching a broad feature set.
At Berkeley, Stanford, and the University of Chicago, among others, students build kernels following the Pintos architecture \cite{pintos}, and at MIT, using Xv6/JOS \cite{xv6}.
In both courses, student kernels support multiprocessor execution;
however, students are also supplied with starter code that already implements thread creation/destruction, context switching, and locking primitives.
In \fourten, we provide starter code for the machine boot sequence and for a minimal C library (\x{printf()}, \x{malloc()}), leaving students to design threading and scheduling primitives from scratch.

Incorporating SMP into the \fourten curriculum would expand the challenges students face in a single semester.
However, we wish to avoid compromising the complexity
%, hardware fidelity,
and learning experience of the project set.
One solution would be to ease the burden of debugging concurrency errors.
Hence, we have built \landslide%
% TODO camready
%\cite{landslide}
, a tool that uses stateless model checking \cite{verisoft,dpor,chess}
as an alternative to the conventional stress tests.
%to force kernel threads to take execution paths that expose concurrency bugs.
\landslide tests student thread libraries fully automatically, and can test kernels provided some manual instrumentation.
It also features an experimental, fully automatic, Pintos testing mode.
We present our experience sharing \landslide with volunteering \fourten students who used it to find bugs in their own code,
and discuss the possibility for integrating it as a main component of operating systems curricula.

>>>>

In this paper we describe our experience teaching concurrent programming at the operating-system level to advanced undergraduate students at \carnegiemellon (\cmu).
In \fourten, the Operating Systems Design and Implementation class at \cmu,
students complete low-level thread library and kernel projects,
which run on real hardware,
in which they implement threads and processes.
Presented with a kernel specification called \pebbles
% TODO camready
\cite{kspec-anonymized}, students build a user-space threading and synchronization library on top of a \pebbles-compliant kernel in a two-week project, and subsequently, in a six-week project, implement their own multi-threaded kernel.
%During these projects
The students face demanding synchronization problems, including how to implement locking primitives and how to coordinate the possibly-simultaneous exiting of parent and child processes.
% Finally, in another two-week project, students implement an extension to their Pebbles kernel. This project changes every semester, and recently has involved supporting symmetric multiprocessing (SMP).

In comparison with other educational kernel projects,
with \pebbles we focus more on ground-up design than on teaching a broad feature set.
At Berkeley, Stanford, and the University of Chicago, among others, students build kernels following the Pintos architecture \cite{pintos}, and at MIT, using Xv6/JOS \cite{xv6}.
In both courses, student kernels support multiprocessor execution;
however, students are also supplied with starter code that already implements thread creation/destruction, context switching, and locking primitives.
In \fourten, we provide starter code for the machine boot sequence and for a minimal C library (\x{printf()}, \x{malloc()}), leaving students to design threading and scheduling primitives from scratch.

Incorporating symmetric multiprocessing (SMP) into the \fourten curriculum would expand the challenges students face in a single semester.
However, we wish to avoid compromising the complexity
%, hardware fidelity,
and learning experience of the project set.
One solution would be to ease the burden of debugging concurrency errors.
Hence, we have built \landslide%
% TODO camready
%\cite{landslide}
, a tool that uses stateless model checking \cite{verisoft,dpor,chess}
as an alternative to the conventional stress tests.
%to force kernel threads to take execution paths that expose concurrency bugs.
\landslide tests student thread libraries fully automatically, and can test kernels provided some manual instrumentation.
It also features an experimental, fully automatic, Pintos testing mode.
We present our experience sharing \landslide with volunteering \fourten students who used it to find bugs in their own code,
and discuss the possibility of integrating it as a main component of the operating systems curriculum.
